Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-29 paper_nips/paper.tex
new figures

Both these biological properties can be conjointly exploited in a mixed
 (i.e., use task and rest data)
low-rank (i.e., perform network decomposition)
approach.
%
The integration of brain-network discovery into
supervised classification formally yields a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
Autoencoders
are one-layered learning models that condense the input data to
local and global representations
by improving reconstruction from them.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
%
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \caption {\textbf{Architecture. XXX develop the cpation. Main msg ?}
  }
\end{wrapfigure}
%
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex transformation \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Hence, autoencoders can flexibly project the neuroimaging data
onto the main axes of variation and thus
reverse-engineer properties of the underlying
neural processes \cite{olshausen96}.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/figure_weights.jpg}
\end{centering}
\caption{\textbf{Comparing the interpretability of classification weights}
The voxel predictors corresponding to 5 (of 18) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (based on same
implementation but without bottleneck layer or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
(latent components $n=20$, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain activity
across the first-level activity maps from 498 participants
(thresholded at 0.25).
All values are z-scored.
% XXX: Why is the z-map sparse ?!? 
}
\end{figure}

@@

@@

\paragraph{Support recovery}
\begin{itemize}
  \item{fit for purpose; the right model should have
  high probability in the right places}
  \item{comparing to ordinary logistic regression, low-rank logreg yielded class weights 
  that were much more similar to feature of the respective training examples,
  for any choice of $\lambda$}
  \item{the higher $\lambda$, the better support recovery}
  \item{conversely, the higher lambda (i.e., bigger emphasis on the
  resting-state autoencoder), the more useful the low-dimensional transformations
  for classification in an independent dataset (ARCHI) as measured by f1 score
  for higher (i.e., 50 and 100) but not lower (i.e., 20) numbers
  of latent components}
\end{itemize}

\paragraph{Anecdotal observations}
\begin{itemize}
  \item{Modifications of the modek that did not improve the generalization performance:
  drop-out, input corruption (p=0.1/0.3/0.5), RELUs}
  \item{no penalty or only L1 or L2 penalty generally yielded worse
  generalization performance}
  \item{low L1/L2 penalties (0.1) yielded better results
  than higher values (0.3, 0.5)}
  \item{``pretraining'' of the bottleneck by settings weights to
  preliminary PCA, SPCA or ICA weights did not give better accuracies,
  neither for autoencoder pretraining }
  \item{present approach better than first PCA, SPCA or ICA and
  subsequently Logreg on loadings, even if each on the whole data}
  \item{introducing a
  nonlinearity (sigmoid, tanh, softplus) into the system
  did not improve predictive accuracy but
  elastic did -> most useful decomposition has characteristics of a SPCA
  and not PCA or ICA}
  \item{introduced an addition layer with many units (50/100) did not
  improve generalization}
\end{itemize}

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of autoencoders, we
learn the optimal decomposition from high-dimensional
voxel brain space into the most important activity patterns for a
supervised learning question.
The higher generalization accuracy and support recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
This domain-specifc classification algorithms
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
neural activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns. That is,
fMRI data probably concentrate near
a low-dimensional manifold of brain networks.
These fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
We hope that these first steps stimulate development towards
powerful semi-supervised classification methods in systems neuroscience.

% grand perspective
In the future, automatic reduction of brain images to
their neurobiological essence
may facilitate data-intense neuroimaging studies.
Initiatives for data collection are rapidly increasing
in neuroscience \cite{poldrack2014data}.
These promise integration
of accumulating neuroscientific knowledge from neuroimaging databases.
Tractability by
condensed feature representations can avoid the ill-posed problem of
learning the full distribution of activity patterns.
%
This is relevant to the 
multi-class challenges spanning the human cognitive space
\cite{schwartz2013mapping}
and
combination of neuroimaging with
high-throughput genetic analyses, such as GWAS
\cite{need2010gwas}.
%
The biggest socioeconomic potential may
lie in across-hospital clinical studies that
predict disease trajectories and drug responses
in psychiatric/neurological populations
\cite{frackowiak2015future, gustav2011cost}.

\paragraph{Acknowledgment}
{\small
Data were provided the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.).
}

>>>>

Both these biological properties can be conjointly exploited in a mixed
 (i.e., use task and rest data)
low-rank (i.e., perform network decomposition)
approach.
%
The integration of brain-network discovery into
supervised classification formally yields a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Architecture.}
  Linear autoencoders find an optimal compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  is thus guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders
are one-layered learning models that condense the input data to
local and global representations
by improving reconstruction from them.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex transformation \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Hence, autoencoders can flexibly project the neuroimaging data
onto the main axes of variation and thus
reverse-engineer properties of the underlying
neural processes \cite{olshausen96}.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/figure_weights.jpg}
\end{centering}
\caption{\textbf{Comparing the interpretability of classification weights}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
(latent components $n=20$, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task (thresholded at 0.25).
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and smoothed by a Gaussian (fwhm=4).
}
\end{figure}

@@

out of sample performance: extract features of general relevance
beyond the supervised problem at hand

support recovery:
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.

@@

\paragraph{Support recovery -> feature identification}
\begin{itemize}
  \item{fit for purpose; the right model should have
  high probability in the right places}
  \item{comparing to ordinary logistic regression, low-rank logreg yielded class weights 
  that were much more similar to feature of the respective training examples,
  for any choice of $\lambda$}
  \item{the higher $\lambda$, the better support recovery}
  \item{conversely, the higher lambda (i.e., bigger emphasis on the
  resting-state autoencoder), the more useful the low-dimensional transformations
  for classification in an independent dataset (ARCHI) as measured by f1 score
  for higher (i.e., 50 and 100) but not lower (i.e., 20) numbers
  of latent components}
\end{itemize}

\paragraph{Anecdotal observations}
\begin{itemize}
  \item{Modifications of the modek that did not improve the generalization performance:
  drop-out, input corruption (p=0.1/0.3/0.5), RELUs}
  \item{no penalty or only L1 or L2 penalty generally yielded worse
  generalization performance}
  \item{``pretraining'' of the bottleneck by settings weights to
  preliminary PCA, SPCA or ICA weights did not give better accuracies,
  neither for autoencoder pretraining }
  \item{present approach better than first PCA, SPCA or ICA and
  subsequently Logreg on loadings, even if each on the whole data}
  \item{introducing a
  nonlinearity (sigmoid, tanh, softplus) into the system
  did not improve predictive accuracy but
  elastic did -> most useful decomposition has characteristics of a SPCA
  and not PCA or ICA}
  \item{introduced an addition layer with many units (50/100) did not
  improve generalization}
\end{itemize}

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of autoencoders, we
learn the optimal decomposition from high-dimensional
voxel brain space into the most important activity patterns for a
supervised learning question.
The higher generalization accuracy and support recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
This domain-specifc classification algorithms
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
neural activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns. That is,
fMRI data probably concentrate near
a low-dimensional manifold of brain networks.
Extracting fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
We hope that these first steps stimulate development towards
powerful semi-supervised classification methods in systems neuroscience.

% grand perspective
In the future, automatic reduction of brain images to
their neurobiological essence
may leverage data-intense neuroimaging studies.
Initiatives for data collection are rapidly increasing
in neuroscience \cite{poldrack2014data}.
These promise structured integration
of accumulating neuroscientific knowledge from neuroimaging databases.
Tractability by
condensed feature representations can avoid the ill-posed problem of
learning the full distribution of activity patterns.
%
This is not only relevant to the 
multi-class challenges spanning the human cognitive space
\cite{schwartz2013mapping}
but also the
multi-modal combination with
high-resolution 3D models of brain anatomy \cite{amunts2013bigbrain}
and
high-throughput genome analyses \cite{need2010gwas}.
%
The biggest socioeconomic potential may
lie in across-hospital clinical studies that
predict disease trajectories and drug responses
in psychiatric/neurological populations
\cite{frackowiak2015future, gustav2011cost}.

\paragraph{Acknowledgment}
{\small
Data were provided by the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.).
}
